Dish: Distribution-oblivious Shell Scripting

### 2019-11-22 pldi/draft.tex

We address this issue by designing a lightweight language whose
purpose is to categorize a command, describes how it reads its input
(meaning in what order), and how to distribute it (in case it is
pure). \kk{I am not sure if the order in which a command reads its
  input will be part of the language since many commands can be made
  to read from one place (like stdin or their first argument). Even in
  that case though, where the command only reads from one argument, we
  have to make sure that there is a way for the user to pinpoint which
  argument does a command read from.}

@@

Consider the command \texttt{cat} \kk{Maybe that is not he best
  example. TODO: Find a good example of a simple command that is in
  different classes depending on the arguments.}. When used without
any flag it is stateless, etc...

@@

%% Can be tricked according to this:
%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars

@@

Sketch of the language:

Some kind of lightweight annotation configuration language (like
yaml), that contains a record for each command.

@@

\kk{Would it be meaningful to give the syntaxx of the language
  formally?}

@@

\kk{Question: Do we need to handle all commands (even the ones that
  could have their input split in several different arguments? or can
  we somehow easily make these ones just take their input from one
  place (e.g. by adding a cat before them).}

\kk{Possible solutions for this (if we have already turned commands
  with more that one input to commands with one input): Have a
  positional reference to the argument that is input (or ``stdin'' if
  it is the stdin). Example: the last argument [-1] of grep is its
  input.}

\kk{It seems that even the arguments might depend on flags (e.g. grep
  checks the current directory if asked recursively, or the stdin
  otherwise), so maybe we should again have a sequence of predicates}

@@

\TODO{Replace texttt with inline listings?}

@@

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output. An important
observation is that commands in the shell might read their input
stream from a sequence of different sources. An illustrative example
is the command \texttt{cat} which reads its input from a combination
of files and its standard input.

@@

The example above reads file \texttt{x} until it encounters the EOF
character, then reads from \texttt{stdin}, and then reads from
\texttt{y}. In order to handle this particularity, nodes in the
dataflow graph can have many incoming edges, that are ordered in a
sequence. Note that this does not correspond to standard dataflow
graph representations where the semantics of multiple incoming edges
is either zipping the inputs in pairs, or arbitraliy interleaving them.

@@

As mentioned above, the dataflow graph exposes task
parallelism. Different nodes in the graph can be executed
concurrently, communicating through the edges. Task level parallelism
is already exposed by the shell, by using pipes \texttt{|} and the
background operator \texttt{\&}. Unfortunately, simple task
parallelism leaves a lot to be desired, since scaling a computation
requires manually introducing more operators.

@@

As mentioned in \Cref{ir}, in order to generate the distributed
implementation of a shell script, \texttt{Dish} first translates it to
a dataflow graph. However, there are several components in shell
programs that cannot be arbitrarily parallelized without affecting the
program behaviour, e.g. commands producing an irreversible effect, or
commands connected with a \texttt{\&\&} operator. In addition, the
fact that the shell is \todo{extremely dynamic}, further increases the
challenge of correctly identifying what parts can be distributed.

@@

Notice that the structure of a shell program clearly identifies
components that can be executed concurrently, and barriers that
enforce synchronization. More precisely, \texttt{cat f1 f2} and
\texttt{grep ``foo''} would already execute concurrently in the
standard shell, while \texttt{sort f3} waits for their completion
before being executed. Both \texttt{cat f1 f2 | grep ``foo'' > f3} and
\texttt{sort f3} are maximal distributable regions, in the sense that
they cannot be extended beyond \texttt{\&\&}.

Concretely, a distributable region cannot permeate through the
following barrier constructs: ;, \&\&, \texttt{||}, function
definitions, command substitutions, and control flow primitives. On
the other hand, the pipe operator \texttt{|} and the parallel
composition operator \& enable parallelization, and extend
distributable regions. Essentially, distributed regions correspond to
subsections of the program that would be executed by different shell
processes.

@@

In the above example\kk{Can we make example not floating?}, while
\texttt{sort f1 > g1}, \texttt{sort f2 > g2}, and \texttt{sort -m g1
  g2} are all in different dataflow graphs. The dataflow graph that
corresponds to the whole program, has two nodes, \texttt{sort f1 > g1
  \&\& sort f2 > g2} and \texttt{sort -m g1 g2}.

@@

Whenever the algorithm encounters a pipe node, it recursively
translates its subcomponents, and then merges them in a dataflow graph
by connecting the output of the first with the input of the second,
\etc. Before the different dataflow graphs are connected, the
algorithm checks that at most one node in the graph writes at every
file, and at most one node reads from a file that another node writes
to. This is important, in order to avoid inconsistent behaviour after
distributing nodes in the dataflow graph \kk{Does this need an
  example?}. Whenever the algorithm encounter an \texttt{\&\&} node,
it recursively translates its subcomponents (similarly to the pipe
node), but then doesn't merge them in a large dataflow graph. Finally,
whenever the algorithm encounters a command, it first checks whether
it is safe to possibly parallelize this command. A command is
considered to be safe, if it does not make any modification to the
shell state (e.g. changing the current directory, modifying
environment variables), and if it does not read from a
\todo{read-once} file, like a fifo pipe. \kk{Mention: Note that an
  unsafe command can still be safely added to the dataflow graph as a
  single node, since this doesn't introduce any new
  behaviours. However it cannot be parallelized further than that.} If
the command is considered safe, its category is identified and a node
containing that command is created and added to a singleton dataflow
graph.

@@

The \texttt{Dish} optimizer is given a dataflow graph representing a
shell program and outputs an optimized dataflow graph, where data
parallelism opportunities have been exploited. More concretely, the
optimizer starts by splitting input files in several chunks, given the
desirable output dataflow graph width. It then starts from the source
nodes of the graph, iteratively performing the graph transformations
that were described in \Cref{ir:transformations}. If the incoming
edges of the node are less than the desirable graph width, it performs
edge splits, and then it performs maximal node splits. The final graph
can then be implemented by spawning a process for each node, and
redirecting the outputs according to the graph edges.

>>>>

We address this issue by designing a lightweight language whose
purpose is to categorize a command, describes how it reads its input
(meaning in what order), and how to distribute it (in case it is
pure).

%% \kk{I am not sure if the order in which a command reads its
%%   input will be part of the language since many commands can be made
%%   to read from one place (like stdin or their first argument). Even in
%%   that case though, where the command only reads from one argument, we
%%   have to make sure that there is a way for the user to pinpoint which
%%   argument does a command read from.}

@@

Consider the command \ttt{cat} \kk{Maybe that is not he best
  example. TODO: Find a good example of a simple command that is in
  different classes depending on the arguments.}. When used without
any flag it is stateless, etc...

@@

Pythonic range for several inputs/outputs

%% Can be tweaked according to this:
%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars

@@

@@

@@

@@

@@

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output. An important
observation is that commands in the shell might read their input
stream from a sequence of different sources. An illustrative example
is the command \ttt{cat} which reads its input from a combination
of files and its standard input.

@@

The example above reads file \ttt{x} until it encounters the EOF
character, then reads from \ttt{stdin}, and then reads from
\ttt{y}. In order to handle this particularity, nodes in the
dataflow graph can have many incoming edges, that are ordered in a
sequence. Note that this does not correspond to standard dataflow
graph representations where the semantics of multiple incoming edges
is either zipping the inputs in pairs, or arbitraliy interleaving them.

@@

As mentioned above, the dataflow graph exposes task
parallelism. Different nodes in the graph can be executed
concurrently, communicating through the edges. Task level parallelism
is already exposed by the shell, by using pipes \ttt{|} and the
background operator \ttt{\&}. Unfortunately, simple task
parallelism leaves a lot to be desired, since scaling a computation
requires manually introducing more operators.

@@

As mentioned in \Cref{ir}, in order to generate the distributed
implementation of a shell script, \ttt{Dish} first translates it to
a dataflow graph. However, there are several components in shell
programs that cannot be arbitrarily parallelized without affecting the
program behaviour, e.g. commands producing an irreversible effect, or
commands connected with a \ttt{\&\&} operator. In addition, the
fact that the shell is \todo{extremely dynamic}, further increases the
challenge of correctly identifying what parts can be distributed.

@@

Notice that the structure of a shell program clearly identifies
components that can be executed concurrently, and barriers that
enforce synchronization. More precisely, \ttt{cat f1 f2} and
\ttt{grep ``foo''} would already execute concurrently in the
standard shell, while \ttt{sort f3} waits for their completion
before being executed. Both \ttt{cat f1 f2 | grep ``foo'' > f3} and
\ttt{sort f3} are maximal distributable regions, in the sense that
they cannot be extended beyond \ttt{\&\&}.

Concretely, a distributable region cannot permeate through the
following barrier constructs: ;, \&\&, \ttt{||}, function
definitions, command substitutions, and control flow primitives. On
the other hand, the pipe operator \ttt{|} and the parallel
composition operator \& enable parallelization, and extend
distributable regions. Essentially, distributed regions correspond to
subsections of the program that would be executed by different shell
processes.

@@

In the above example\kk{Can we make example not floating?}, while
\ttt{sort f1 > g1}, \ttt{sort f2 > g2}, and \ttt{sort -m g1
  g2} are all in different dataflow graphs. The dataflow graph that
corresponds to the whole program, has two nodes, \ttt{sort f1 > g1
  \&\& sort f2 > g2} and \ttt{sort -m g1 g2}.

@@

Whenever the algorithm encounters a pipe node, it recursively
translates its subcomponents, and then merges them in a dataflow graph
by connecting the output of the first with the input of the second,
\etc. Before the different dataflow graphs are connected, the
algorithm checks that at most one node in the graph writes at every
file, and at most one node reads from a file that another node writes
to. This is important, in order to avoid inconsistent behaviour after
distributing nodes in the dataflow graph \kk{Does this need an
  example?}. Whenever the algorithm encounter an \ttt{\&\&} node,
it recursively translates its subcomponents (similarly to the pipe
node), but then doesn't merge them in a large dataflow graph. Finally,
whenever the algorithm encounters a command, it first checks whether
it is safe to possibly parallelize this command. A command is
considered to be safe, if it does not make any modification to the
shell state (e.g. changing the current directory, modifying
environment variables), and if it does not read from a
\todo{read-once} file, like a fifo pipe. \kk{Mention: Note that an
  unsafe command can still be safely added to the dataflow graph as a
  single node, since this doesn't introduce any new
  behaviours. However it cannot be parallelized further than that.} If
the command is considered safe, its category is identified and a node
containing that command is created and added to a singleton dataflow
graph.

@@

The \ttt{Dish} optimizer is given a dataflow graph representing a
shell program and outputs an optimized dataflow graph, where data
parallelism opportunities have been exploited. More concretely, the
optimizer starts by splitting input files in several chunks, given the
desirable output dataflow graph width. It then starts from the source
nodes of the graph, iteratively performing the graph transformations
that were described in \Cref{ir:transformations}. If the incoming
edges of the node are less than the desirable graph width, it performs
edge splits, and then it performs maximal node splits. The final graph
can then be implemented by spawning a process for each node, and
redirecting the outputs according to the graph edges.
